---
title: "Analysis of temporal trends in southern resident killer whales"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: true
  pdf_document:
    toc: yes
    toc_depth: '3'
---

***

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

***

```{r set_options, echo = FALSE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
```

# Requirements

All analyses require the [R software](https://cran.r-project.org/) (v3.4+) for data retrieval, data processing, and summarizing model results. We also need a few packages that are not included with the base installation of R, so we begin by installing them (if necessary) and then loading them.

```{r load_pkgs, message = FALSE, warning = FALSE}
if(!require("here")) {
  install.packages("here")
  library("here")
}
if(!require("readr")) {
  install.packages("readr")
  library("readr")
}
if(!require("ggplot2")) {
  install.packages("ggplot2")
  library("ggplot2")
}
## set directory locations
datadir <- here("data")
analdir <- here("analysis")
```

# Data

The counts of southern resident killer whales (SRKW) come from the [Center for Whale Research](https://www.whaleresearch.com/orcasurvey), and are broken down by the three major groups: J, K and L pods.

```{r read_data}
dat <- read_csv(file.path(datadir, "SRKW.csv"))
head(dat)
```

