---
title: "Analysis of temporal trends in southern resident killer whales"
output:
  html_document:
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: true
  pdf_document:
    toc: yes
    toc_depth: '3'
---

***

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

***

```{r set_options, echo = FALSE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
```

# Requirements

All analyses require the [R software](https://cran.r-project.org/) (v3.4+) for data retrieval, data processing, and summarizing model results. We also need a few packages that are not included with the base installation of R, so we begin by installing them (if necessary) and then loading them.

```{r load_pkgs, message = FALSE, warning = FALSE}
if(!require("here")) {
  install.packages("here")
  library("here")
}
if(!require("readr")) {
  install.packages("readr")
  library("readr")
}
if(!require("ggplot2")) {
  install.packages("ggplot2")
  library("ggplot2")
}
## set directory locations
datadir <- here("data")
analdir <- here("analysis")
```

# Data

## Southern resident killer whales

The counts of southern resident killer whales (SRKW) come from the [Center for Whale Research](https://www.whaleresearch.com/orcasurvey), and are broken down by the three major groups: J, K and L pods.

```{r read_srkw}
kw <- read_csv(file.path(datadir, "SRKW.csv"))
head(kw)
n_yr <- nrow(kw)
```

## Snake River spring Chinook

The counts of Snake River spring Chinook at Lower Granite Dam come from [Columbia Basin Research](http://www.cbr.washington.edu/dart/query/adult_annual_sum). I'll divide them by 1000 to make them more similar in magnitude to the SRKW data.

```{r get_srss}
ss <- read_csv(file.path(datadir, "SRSS.csv"))
ss[,2] <- ss[,2]/1000
head(ss)
```

# Plots

Here are plots of the time series of counts for orcas and salmon.

```{r plot_counts, fig.height=8, fig.width=6}
par(mfrow=c(2,1), mai=c(0.8,0.8,0.2,0), omi=rep(0.1,4))
clr <- rev(viridis::viridis(4, end = 0.8))
## orcas
matplot(kw[,1], kw[,-1], type="l", lty="solid", xlim = c(kw[[1,1]], 2018),
        xlab = "", ylab = "Count", col = clr)
text(2018,kw[nrow(kw),-1],c("J","K","L","All"), cex=0.9, col = clr)
mtext("Southern resident killer whales", 3, adj=0, line=0.5)
## salmon
matplot(ss[,1], ss[,2], type="l", lty="solid", xlim = c(kw[[1,1]], 2018),
     xlab = "Year", ylab = "Count (1000s)", col = viridis::viridis(1))
mtext("Snake River spring Chinook", 3, adj=0, line=0.5)
```

It is perhaps more informative to look at the year-to-year differences in the counts of SRKW, as that gives an indication as to whether births exceed deaths.

```{r plot_diffs, fig.height=4, fig.width=6}
## difference ts
kwd <- apply(kw[,2:4], 2, diff)
par(mai=c(0.8,0.8,0.2,0), omi=rep(0.1,4))
matplot(kw[-1,1], kwd, type="l", lty="solid",
        xlab = "", ylab = "Count", col = clr)
```

# Correlations

Here are the correlations among the differenced counts of SRKW and SRSS Chinook.

```{r corr}
## raw counts of Chinook
round(cor(x = ss[-n_yr,2], y = kwd), 2)
## log counts of Chinook
round(cor(x = log(ss[-n_yr,2]), y = kwd), 2)
```
